A novel method for detecting coalescing binaries in near realtime with Advanced LIGO and beyond \editorial{Submit to PRD or CQG?}

### 2011-02-04 analysis.tex
Huge set of changes
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@598 ccdd05d5-56c1-461a-bf98-85f6876523b4

@@

>>>>

\editorial{Should I mention \texttt{GStreamer} at all?} The detection method described above is much more efficient in terms of floating point operations than the traditional matched filter bank method.  However, time slices and conditional reconstruction greatly complicate queueing, synchronizing, and bookkeeping of intermediate signals.  A low latency implementation capable of recruiting more than one \textsc{cpu} core would be difficult to achieve within the familiar serial programming framework because of the nontrivial time-delay relationships between samples.  Due to these complications, we chose to prototype the search using an open source signal processing environment called \texttt{GStreamer}.  \editorial{Citation for \texttt{GStreamer}?}  Primarily used for playing, authoring, or streaming media on \texttt{Linux} systems, \texttt{GStreamer} is an integral component of the popular \texttt{Gnome} desktop.

\editorial{It would be good to illustrate the layout of this particular template bank: masses spanned, time slice layout \dots}
\begin{table}[!h]
\caption{Operation counts per sample for six different detection methods.  The operation counts for \textsc{lloid} assume a reconstruction duty cycle of 5\%.  Note that the \textsc{fir} method with \textsc{lloid} is almost 10 times faster than the conventional \textsc{fft} method, despite having substantially lower latency.}
\begin{center}
\input{flop_budget}
\end{center}
\end{table}

\begin{comment}

% Ugly diagram!

@@

\end{comment}

### 2011-02-04 appendix.tex
Huge set of changes
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@598 ccdd05d5-56c1-461a-bf98-85f6876523b4

>>>>

\section{Floating point operation counts}

\editorial{I just yanked this from the methods section.  It needs to be made more concise, maybe turned into a table of CPU budgets for various components of LLOID.} The filter bank can be implemented using finite impulse response (\textsc{fir}) filters, which are just sliding window dot products.  If there are $M$ templates of length $n$, and the data stream contains $N$ samples, then applying the filter bank requires $2 M N n$ operations.

More commonly, the matched filters are implemented using the \textsc{fft} convolution.  This entails applying \textsc{fft}s to blocks of $D$ samples, with $2 n \leq D$, each block overlapping the previous one by $n$ samples.  There are $N/(D-n)$ such blocks.  Modern implementations of the Cooley-Tukey \textsc{fft}, such as the ubiquitous \texttt{fftw}, require about $4 N \lg N$ operations to evaluate a \textsc{dft} of size $N$~\cite{Johnson:2007p9654}.  \editorial{This is more commonly known as ``overlap-save''.  We should find someone else's operation count and cite it.}  A $D$ sample cross-correlation consists of a forward \textsc{fft}, an $D$ sample dot product, and an inverse \textsc{fft}, totaling $8 D \lg N + 2 D$ operations per block.  Per sample, this is $(8 \lg D + 2) / (1 - n/D)$ operations.

The \textsc{fir} filter implementation has the advantage that it has no intrinsic latency, whereas the \textsc{fft} convolution has at least the latency of the \textsc{fft} block size $D \geq 2 n$.  For example, for a $1.4 - 1.4 \, M_\odot$ template with duration $\sim 1 \, \mathrm{ks}$, the \textsc{fft} convolution has a latency $\geq 2 \, \mathrm{ks}$.  However, the \textsc{fir} filter implementation has the disadvantage of much greater overhead per sample than the \textsc{fft} convolution.  For a $1\,\mathrm{ks}$ template sampled at $4096\,\mathrm{Hz}$, the \textsc{fir} implementation requires about about $n / 8 \lg 2 n = 2.2 \times 10^4$ times more operations per sample than the \textsc{fft} implementation.

\begin{comment}
% Let's make glitch rejection in LLOID a separate paper.

### 2011-02-04 inspiral_svd.tex
Huge set of changes
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@598 ccdd05d5-56c1-461a-bf98-85f6876523b4

\documentclass[aps,prd,twocolumn,showpacs,groupedaddress,showkeys,preprintnumbers]{revtex4}
\input{packages}
\input{macros}

% $Id:$

@@

\title{Low-latency search for gravitational-waves: prospects for near real time detection of compact binary coalescence}

@@

\author{Kipp Cannon} 
% \email{antony.searle@anu.edu.au}
\affiliation{The LIGO Laboratory - California Institute of Technology, Pasadena, CA 91125}

\author{Adrian Chapman} 
% \email{antony.searle@anu.edu.au}
\affiliation{The LIGO Laboratory - California Institute of Technology, Pasadena, CA 91125}

\author{Chad Hanna}  
% \email{antony.searle@anu.edu.au}
\affiliation{The LIGO Laboratory - California Institute of Technology, Pasadena, CA 91125}

\author{Drew Keppel}  
% \email{antony.searle@anu.edu.au}
\affiliation{The LIGO Laboratory - California Institute of Technology, Pasadena, CA 91125}

\author{Antony C. Searle}
% \email{antony.searle@anu.edu.au}
\affiliation{The LIGO Laboratory - California Institute of Technology, Pasadena, CA 91125}

\author{Alan J. Weinstein}
% \email{antony.searle@anu.edu.au}
\affiliation{The LIGO Laboratory - California Institute of Technology, Pasadena, CA 91125}

\begin{abstract}

\end{abstract}

\pacs{}

\keywords{Gravitational Waves, Laser Interferometry}

\preprint{}

\maketitle

\input{introduction}

@@

\begin{acknowledgments}
LIGO was constructed by the California Institute of Technology and Massachusetts Institute of Technology with funding from the National Science Foundation and operates under cooperative agreement PHY-0107417. This paper has LIGO Document Number LIGO-P0900004-v1

\end{acknowledgments}

\appendix

\bibliography{references}

>>>>

\documentclass[showpacs,groupedaddress,showkeys,preprintnumbers]{iopart}
\input{packages}
\input{macros}

@@

\title[LLOID]{A novel method for detecting coalescing binaries in near realtime with Advanced LIGO and beyond \editorial{Submit to PRD or CQG?}}

@@

\author{Kipp Cannon$^1$, Adrian Chapman$^2$, Chad Hanna$^3$, Drew Keppel$^4$, Antony C. Searle$^2$, Alan J. Weinstein$^2$}

\address{$^1$ Canadian Institute for Theoretical Astrophysics, ON, Canada}
\address{$^2$ LIGO Laboratory - California Institute of Technology, Pasadena, CA, USA}
\address{$^3$ Perimeter Institute for Theoretical Physics, Waterloo, ON, Canada}
\address{$^3$ Albert-Einstein-Institut, Max-Planck-Institut f\"{u}r Gravitationphysik, Hannover, Germany}

\begin{abstract}
Conventional matched filter bank methods for the detection of gravitational waves from the inspiral of compact binaries are computationally expensive, have hundreds of seconds of unavoidable intrinsic latency, and require arrays of largely redundant matched filters.  Novel detection methods that are more computationally efficient and have lower latency will be required to realize the full potential of advanced of gravitational wave detectors that are currently under construction.  In this paper, we describe a new detection method that exploits the properties of inspiral waveforms using multi-rate filtering, principal component analysis, and hierarchical detection.  We provide receiver operating characteristics from a prototype search pipeline that is capable of low-latency or near-realtime detection with greatly reduced computational requirements in comparison with previously described methods.
\end{abstract}

\pacs{95.55.Ym, 84.30.Vn, 95.75.Wx}

%\keywords{Gravitational Waves, Laser Interferometry}

%\preprint{}
\input{introduction}

@@

\ack{\textsc{ligo} was constructed by the California Institute of Technology and Massachusetts Institute of Technology with funding from the National Science Foundation and operates under cooperative agreement \textsc{phy}-\oldstylenums{0107417}. This paper has \textsc{ligo} Document Number \textsc{ligo}-\textsc{p}\oldstylenums{0900004}-v\oldstylenums{1}.

\appendix
\input{appendix}

\section*{References}

\bibliographystyle{unsrt}
\bibliography{references}

### 2011-02-04 introduction.tex
Huge set of changes
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@598 ccdd05d5-56c1-461a-bf98-85f6876523b4

\section{Introduction}
\label{sec:introduction}
Modern gravitational-wave observatories are likely to deliver on the promise 
that their name implies by observing multiple gravitational-wave signals from
compact binary coalescence within the coming decade ~\cite{kopparapu2008}. 
There is now a world-wide network of gravitational-wave observatories 
including the laser interferometers
LIGO, VIRGO, GEO, and TAMA, which should assure that coincident detections 
will be possible among at least three separate sites lending confidence
to any detections.
~\cite{Abbott:2005qm, beauville2008, beauville2006, blackburn2005, LIGOS1instpaper}.  
However, having an electromagnetic or neutrino counterpart to a 
gravitational-wave detection would not only increase the confidence in 
the detection but would also greatly improve the astrophysical information
available from the event.

Work has already commenced to trigger gravitational-wave searches from 
electromagnetic observations ~\cite{triggeredsearches2008}. LIGO has recently
ruled out the possibility of GRB070201 originating from the merger of a
neutron star / neutron star or neutron star / black hole system in the
Andromeda galaxy ~\cite{GRB070201}.  The still open question is, will 
gravitational-wave observations be able to trigger electromagnetic followup?

This question has been investigated and it seems  at least
possible that gravitational-wave detections could provide a region of the
sky to prompt electromagnetic observation followups~\cite{sylvestre2003}.
Much work is currently underway in providing the best possible source 
localization from gravitational-wave detector networks
~\cite{markowitz2008, raymond2008, cavalier2006}.  Colloborations are
currently forming to provide infrastructure for the gravitational-wave
Astronomers to provide targets of opportunity for electromagnetic 
Astronomers ~\cite{kanner2008}.  Most models that predict simultaneous
gravitational-wave and electromagnetic observations also predict that the
peak amplitude of electromagetic radiation will occur soon after 
gravitational-wave emmission ~\cite{sylvestre2003}.  Thus in order to maximize
the chance of a successful electromagnetic followup the latencey of 
gravitational-wave signal analysis must be made minimal.

The parameter space of compact binary coalescence signals is large
~\cite{Owen:1995tm, Owen:1998dk} .  It is a
computationally burdensome task to analyze these signals with even moderate
latency ~\cite{Abbott:2007xi}.  
This work will describe how to exploit degeneracy in the signal
parameter space to answer more quickly whether or not a gravitational-wave 
is present, with what confidence, and where it is coming from. Specifically
we will explore using the singular value decomposition (SVD) to reduce the
effective number of filters required to search the data.  We note that others
have applied the use of SVD to gravitational wave data analysis to analyze
optimal gravitational-wave burst detection 
~\cite{bradyraymajumder2004, heng2008} and coherent networks 
of detectors ~\cite{wen2008}

>>>>

\section{Introduction}
\label{sec:introduction}

% I would like to say:
%  - Low mass inspirals are the most promising source of gravitational radiation for advanced LIGO
% - Mention S6 and VSR2, online running, EM followup
% - Observational applications of near-realtime detection
% - Point out challenges of advanced detectors: more templates, longer templates
% - Advanced LIGO timetable
% - Einstein telescope
% - Layout of paper: description of method, procedure followed

\begin{comment}

However, the computational requirements of conventional detection strategies --- especially those that are well suited to low latency applications --- will be greatly increased by this upgrade for two reasons.  First, advanced detectors will have much improved low-frequency sensitivity, so chirp signals from coalescing binaries will remain in the detectors' band for longer.  This will demand that matched filter based searches use longer templates, requiring more memory and more cycles to process.  Second, background noise for advanced detectors will be more spectrally uniform, so it will be possible to resolve the intrinsic parameters of a source that determine the time-frequency structure of the gravitational wave signal with much greater accuracy.  This will require that matched filter bank based searches employ more matched filters, again demanding more memory and more cycles.

\end{comment}

\begin{comment}

\editorial{This paragraph has too many citations.} Modern gravitational-wave observatories are likely to deliver on the promise 
that their name implies by observing multiple gravitational-wave signals from
compact binary coalescence within the coming decade \cite{kopparapu2008}. 
There is now a world-wide network of gravitational-wave observatories 
including the laser interferometers
\textsc{ligo}, Virgo, \textsc{geo}, and \textsc{tama}, which should assure that coincident detections 
will be possible among at least three separate sites lending confidence
to any detections~\cite{Abbott:2005qm, beauville2008, beauville2006, blackburn2005, LIGOS1instpaper}.  \editorial{This is a lot of citations just to say that the 4 detectors exist.  Maybe if they were scattered throughout the paragraph?}
However, having an electromagnetic or neutrino counterpart to a 
gravitational-wave detection would not only increase the confidence in 
the detection but would also greatly improve the astrophysical information
available from the event.

\end{comment}

In a series of detector upgrades that are now under way, \textsc{ligo} and Virgo will become the first advanced gravitational wave detectors, gaining a tenfold improvement in amplitude sensitivity \citeneeded{} and a corresponding thousandfold increase in observable volume in the local universe.  It is anticipated that these detectors will be able to sense as many as 40 events per year \cite{Abadie:2010p10836}.

As the gravitational wave detection horizon pushes outward, the ability to detect signals in near realtime will become valuable.  Having an electromagnetic or neutrino counterpart to a 
gravitational-wave detection would not only increase the confidence in 
the detection but will also greatly improve the astrophysical information
available from the event.  Most models that predict simultaneous
gravitational-wave and electromagnetic observations also predict that the
peak amplitude of electromagetic radiation will occur soon after 
gravitational-wave emmission~\cite{sylvestre2003}.  Thus in order to maximize
the chance of a successful electromagnetic followup the latency of 
gravitational-wave signal analysis must be made minimal.

\editorial{This is old news, and doesn't seem relevant to me: EM-directed gravitational wave searches don't benefit from low-latency detection pipelines.} Work has already commenced to trigger gravitational-wave searches from 
electromagnetic observations~\cite{triggeredsearches2008}. \textsc{ligo} observations
ruled out the possibility of GRB70201 originating from the merger of a
\textsc{nsns} or \textsc{nsbh} system in the Andromeda galaxy~\cite{GRB070201}.

In \oldstylenums{2010}, \textsc{ligo} and Virgo completed a period of joint data taking during which several all-sky detection pipelines operated in a low-latency configuration \citeneeded.  Promising detection candidates were promptly sent for electromagnetic followup \citeneeded{} to several telescopes including Swift, \textsc{rotse}, \textsc{tarot}, and Zadko.

This question has been investigated and it seems  at least
possible that gravitational-wave detections could provide a region of the
sky to prompt electromagnetic observation followups~\cite{sylvestre2003}.
\editorial{Cite an in-preparation paper of Larry's?} Much work is currently underway in providing the best possible source 
localization from gravitational-wave detector networks~\cite{markowitz2008, raymond2008, cavalier2006}.  Colloborations are currently forming to provide infrastructure for the gravitational-wave
Astronomers to provide targets of opportunity for electromagnetic 
astronomers~\cite{kanner2008}.  Most models that predict coincident
gravitational-wave and electromagnetic observations also predict that the
peak electromagetic fluence will occur soon after 
gravitational-wave emission~\cite{sylvestre2003}.  Thus in order to maximize
the chance of a successful electromagnetic followup the latency of 
gravitational-wave signal analysis must be made minimal.

The parameter space of compact binary coalescence signals is large~\cite{Owen:1995tm, Owen:1998dk}.  It is a
computationally burdensome task to analyze these signals with even moderate
latency~\cite{Abbott:2007xi}.  
This work will describe how to exploit degeneracy in the signal
parameter space to answer more quickly whether or not a gravitational-wave 
is present. Specifically we will explore using the singular value decomposition (\textsc{svd}) to reduce the
effective number of filters required to search the data.  We note that others
have applied the use of \textsc{svd} to gravitational wave data analysis to analyze
optimal gravitational-wave burst detection~\cite{bradyraymajumder2004, heng2008} and coherent networks 
of detectors~\cite{wen2008} .
